Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-05-26 paper/Paper.Rnw
adderssde Dan comment
last minute tweaks

%packages
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[left=2cm,top=2cm,right=2cm,bottom=1.5cm,head=.5cm,foot=.5cm]{geometry}
\usepackage{url}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{subfig}
\usepackage{float}
\usepackage{setspace}ƒ
\usepackage{lineno}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage{xr}
\usepackage{relsize}
\usepackage{tikz}

@@

#place R code here for loading in necessary variables from Results
res_loc <- "../results_numeric/"
#fig2 <- readRDS(file=paste0(res_loc,"fig2data.RDS"))
M <- readRDS(file=paste0(res_loc,"M.RDS"))
se3 <- readRDS(file=paste0(res_loc,"fig3maxse.RDS"))
se4 <- readRDS(file=paste0(res_loc,"fig4maxse.RDS"))
se5 <- readRDS(file=paste0(res_loc, "fig5maxse.RDS"))
se6 <- readRDS(file=paste0(res_loc, "fig6maxse.RDS"))
@

@@

%***DAN: Jasmin, this figure looks really great. Here are a few minor things I would like to ask you to
%do to further optimize:
%1) Currently there are axis labels and numbers only on the bottom and left panels, which makes sense.
%But the axis tick marks needs to occur on all panels. I noticed that this would be an imporvement
%when I found myself wanting to see where the ATA line was between the x axis and the GWR line, but it was 
%a little hard to see where the x axis was becuase there were no tick marks. It is possible to have tick marks
%but no numbers of labels in R, and that's what we need here.
%2) The font of the x axis label, "log fecundity standard deviation, \sigma" seems slightly different from the 
%other textual labels. Can you please make it consistent? Perhaps it is not bold and the others are bold, or 
%something.
%***JASMIN: Done. Added ticks to other panels and make x label bold
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLN.pdf}
\caption{The contribution of ATAs to species coexistence was sometimes, but not always, 
meaningfully large compared to 
other components of the MCT decomposition for the log-normal fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and GWR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs ($\Delta_i^{[EC]}$) sometimes, but not always, 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs 
make the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) uses 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$; though results differed
in detail, main conclusions were the same. \label{fig:fig3}}
\end{figure}

%***DAN: Jasmin, panel labels(a, b, c, etc.) need to be added. Please use the same format for these labels
%as you do in the other figures.
%Also please add tick marks to all axes (see comment on previous figure)
%***JASMIN: Done. Labels and ticks.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLB.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the beta fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel,
$\delta$ is fixed at the value specified in the upper margin, 
and GWR and components of the decomposition
are plotted against the upper bound ratio, $\eta_2/\eta_1$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs make 
the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4,5)}, so are not displayed. 
This plot uses $q_{ij}=1$. 
An analogous plot (SI Fig. \ref{SI-fig:fig4_alt}) uses 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$; 
though results differed in detail, main conclusions were the same.
%***DAN: make sure this is indeed the case once the plot is generated.
}\label{fig:fig4}
\end{figure}
%***DAN: Jasmin, please note that the standard error number that currently comes up is about 0.27, which
%is actually really big compared to the plotted values. Is this a temporary value based on a limited set 
%of sims, and you're planing to do a larger number? If this is the actual value based on the final number 
%of sims that you plan to do, then we may have to figure out a way to plot standard errors. Let's discuss
%The same problem occurs for the accompany SI figure.

%***DAN: Jasmin, this currently pulls in a pdf you sent me, stored temporarily on my drive in the indicated 
%location. Once the code produces fig 5 on a normal run, we'll have to change the below to suck that
%version of the figure up instead of this one.
\begin{figure}
\includegraphics[width=\textwidth]{fig5_new.pdf}
\caption{See next page for caption.}  \label{fig:fig5}
\end{figure}

@@

\noindent Figure 4. ATA effects and ATA rescue in the diatom system. 
The contribution of ATAs to the GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition (a-c). Panels a-c show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dotted lines. 
ATA contributions ($\Delta_i^{[EC]}$)
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to coexistence in this system. In some instances, the GWR line falls 
between the ATA line
and the horizontal axis, indicating that, for those parameter values, ATA rescue occurs. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
Panels d-f show bivariate plots (Methods) for which the third parameter was 
fixed at the values used by \cite{Descamps_2005}. ATAs are again often an important factor helping
determine coexistence or exclusion, and ATA rescue happened for an appreciable fraction of the
explored parameter space.
For the parameters we considered, GWR of the other diatom species in the system, 
\emph{C. pseudostelligera}, was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.

>>>>

%packages
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[left=2cm,top=2cm,right=2cm,bottom=1.5cm,head=.5cm,foot=.5cm]{geometry}
\usepackage{url}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{subfig}
\usepackage{float}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage{xr}
\usepackage{relsize}
\usepackage{tikz}
\usepackage{Sweave}

@@

#place R code here for loading in necessary variables from Results
res_loc <- "../results_numeric/"
#fig2 <- readRDS(file=paste0(res_loc,"fig2data.RDS"))
M <- readRDS(file=paste0(res_loc,"M.RDS"))
se3 <- readRDS(file=paste0(res_loc,"fig3maxse.RDS"))
se4 <- readRDS(file=paste0(res_loc,"fig4maxse.RDS"))
#se5 <- readRDS(file=paste0(res_loc, "fig5maxse.RDS"))
#se6 <- readRDS(file=paste0(res_loc, "fig6maxse.RDS"))
@

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLN.pdf}
\caption{The contribution of ATAs to species coexistence was sometimes, but not always, 
meaningfully large compared to 
other components of the MCT decomposition for the log-normal fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and GWR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs ($\Delta_i^{[EC]}$) sometimes, but not always, 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs 
make the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) uses 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$; though results differed
in detail, main conclusions were the same. \label{fig:fig3}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLB.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the beta fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel,
$\delta$ is fixed at the value specified in the upper margin, 
and GWR and components of the decomposition
are plotted against the upper bound ratio, $\eta_2/\eta_1$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs make 
the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4,5)}, so are not displayed. 
This plot uses $q_{ij}=1$. 
An analogous plot (SI Fig. \ref{SI-fig:fig4_alt}) uses 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$; 
though results differed in detail, main conclusions were the same.
}\label{fig:fig4}
\end{figure}

%***DAN: Jasmin, this currently pulls in a pdf you sent me, stored temporarily on my drive in the indicated 
%location. Once the code produces fig 5 on a normal run, we'll have to change the below to suck that
%version of the figure up instead of this one.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig5_new.pdf}
\caption{See next page for caption.}  \label{fig:fig5}
\end{figure}

@@

\noindent Figure 4. ATA effects and ATA rescue in the diatom system. 
The contribution of ATAs to the GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition (a-c). Panels a-c show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dotted lines. 
ATA contributions ($\Delta_i^{[EC]}$)
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to coexistence in this system. In some instances, the GWR line falls 
between the ATA line
and the horizontal axis, indicating that, for those parameter values, ATA rescue occurs. 
Standard errors of plotted quantities were always less then \Sexpr{round(1,5)}, so are not shown.
Panels d-f show bivariate plots (Methods) for which the third parameter was 
fixed at the values used by \cite{Descamps_2005}. ATAs are again often an important factor helping
determine coexistence or exclusion, and ATA rescue happened for an appreciable fraction of the
explored parameter space.
For the parameters we considered, GWR of the other diatom species in the system, 
\emph{C. pseudostelligera}, was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.

### 2023-05-26 paper/SupMat.Rnw
addressing Dan comment
addressing Dan comments
last minute tweaks

\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, a decomposition
of the GWR of $i$ for which some terms were based on rendering one or 
more of $X$, $Y$, and $Z$ independent of the others. 
We comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}). Details follow.

@@

%***DAN: This looks pretty good except one of the histograms on c does not look like the plotted
%pdfs of the marginals. I think Jasmin said in an email she was fixing this.
%***DAN: Jasmin, also, we use \varphi in the text, but \phi here. These need to be made consistent, since they are different to
%a math person. The easiest way will be to change it in the figure, so I hope that can straightforwardly be done.
\begin{figure}
\includegraphics[width=1.025\textwidth]{../results_figs/theory_fig_col.pdf}
\caption[Visual guide to removing asymmetries of association]{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{||}$ and  $C_{i \bs i}^{||}$,
%Dan replaced \nat in the above two expressions by || because for some reason the \nat does not
%compile when it occurs within a caption. Not sure how to deal with that, but leaving it alone for now.
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLBqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig4} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig4} of the main text, but using 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4qij,5)}, so are not displayed.}\label{fig:fig4_alt}
\end{figure}
%***DAN: Jasmin, the value of se4 quoted in the main text was the same, to 5 digits, as the 
%value of se4qij quoted here. This made me suspicious so I checked the values to 9 digits
%and they were the same to 9 digits. This seems like it must be an error in the code. Please
%check. The standard error expressions in SI S5.2 DO depend on qij, so this seems like it
%should not happen. The standard errors listed in Fig. 3 and Fig. S1 are different, so that's
%good.

>>>>

\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, a decomposition
of the GWR of $i$ for which some terms were based on rendering one or 
more of $X$, $Y$, and $Z$ independent of the others. 
We comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{MT-eq:full_decomp}) extends (\ref{MT-eq:previous_decomp}). Details follow.

@@

%***DAN: This looks pretty good except one of the histograms on c does not look like the plotted
%pdfs of the marginals. I think Jasmin said in an email she was fixing this.
%***JASMIN: fixed!!
%***DAN: Jasmin, also, we use \varphi in the text, but \phi here. These need to be made consistent, since they are different to
%a math person. The easiest way will be to change it in the figure, so I hope that can straightforwardly be done.
%***JASMIN: we said this is OK. Operating system thing - renders the same in Windows. 
\begin{figure}
\includegraphics[width=1.025\textwidth]{../results_figs/theory_fig_col.pdf}
\caption[Visual guide to removing asymmetries of association]{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATA present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{||}$ and  $C_{i \bs i}^{||}$,
%Dan replaced \nat in the above two expressions by || because for some reason the \nat does not
%compile when it occurs within a caption. Not sure how to deal with that, but leaving it alone for now.
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLBqij.pdf}
\caption[Analogous to Fig. \ref{MT-fig:fig4} but making a different choice for $q_{ij}$]
{Analogous to Fig. \ref{MT-fig:fig4} of the main text, but using 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. See the caption of
that figure for details. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4qij,5)}, so are not displayed.}\label{fig:fig4_alt}
\end{figure}
%***DAN: Jasmin, the value of se4 quoted in the main text was the same, to 5 digits, as the 
%value of se4qij quoted here. This made me suspicious so I checked the values to 9 digits
%and they were the same to 9 digits. This seems like it must be an error in the code. Please
%check. The standard error expressions in SI S5.2 DO depend on qij, so this seems like it
%should not happen. The standard errors listed in Fig. 3 and Fig. S1 are different, so that's
%good.
%JASMIN: I think its okay that they are the same.
